Fast processing of JUNGFRAU data with the Alpaka library

### 2020-01-06 Publication/manuscript.tex
Added new bullet points to implementation section of the paper.

\documentclass[a4paper]{article}
\usepackage{authblk}
\usepackage[backend=bibtex]{biblatex}
\usepackage{hyperref}
\usepackage{txfonts}
\usepackage{titling}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[a4paper, margin={2.5cm, 1.7cm}]{geometry}
\usepackage{mwe}
\usepackage{filecontents}
\usepackage{changes}
\usepackage{lipsum}% <- For dummy text
\definechangesauthor[name={Heide}, color=blue]{Heide}
\setremarkmarkup{(#2)}

@@

{\bf Keywords:} Photon pixel detector, fast data processing, GPU programming, Alpaka...

@@

The increasing data rates during FEL experiments require dedicated detectors as well as advanced methods for fast data processing.
At PSI, the Jungfrau detector (\cite{JFapplications}, \cite{JFcalibration}, \cite{JFoperation}, \cite{Mozzanica_2016}) was developed which is capable due to a gain switching scheme to register single photons as well as photon bunches in a charge integration mode. A conversion of the detector data to the energy and the number of registered photons already during data aquisition is beneficial for the analysis of the measurement and furthermore for the reduction of storage space. This online data conversion algorithm has to correct each pixel value using the gain maps and the continously updated pedestal maps following a lab-based calibration  procedure~\cite{JFcalibration}. The summation of a couple of frames and the cluster identification in the data maps further reduce the amount of data. The different modules of one detector system can be processed fully parallel, however, the data conversion itself is not data parallel. Therefore, and due to the need for an advanced parallel implementation which is independent of the computing hardware, an Alpaka~\cite{Matthes17} version was written and its performance was tested.\\

@@

@@

>>>>

\documentclass[a4paper]{article}
\usepackage{authblk}
\usepackage[backend=bibtex]{biblatex}
\usepackage{hyperref}
\usepackage{txfonts}
\usepackage{titling}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[a4paper, margin={2.5cm, 1.7cm}]{geometry}
\usepackage{mwe}
\usepackage{filecontents}
\usepackage{changes}
\usepackage{lipsum}% <- For dummy text
%\definechangesauthor[name={Heide}, color=blue]{Heide}
%\definechangesauthor[name={Jonas}, color=magenta]{Jonas}
%\setremarkmarkup{(#2)}

@@

{\bf Keywords:} Photon pixel detector, fast data processing, GPU programming, Alpaka \textcolor{magenta}{...}

@@

The increasing data rates during FEL experiments require dedicated detectors as well as advanced methods for fast data processing.
At PSI, the Jungfrau detector (\cite{JFapplications}, \cite{JFcalibration}, \cite{JFoperation}, \cite{Mozzanica_2016}) was developed which is capable due to a gain switching scheme to register single photons as well as photon bunches in a charge integration mode. A conversion of the detector data to the energy and the number of registered photons already during data acquisition is beneficial for the analysis of the measurement and furthermore for the reduction of storage space. This online data conversion algorithm has to correct each pixel value using the gain maps and the continuously updated pedestal maps following a lab-based calibration  procedure~\cite{JFcalibration}. The summation of a couple of frames and the cluster identification in the data maps further reduce the amount of data. The different modules of one detector system can be processed fully parallel, however, the data conversion itself is not data parallel. Therefore, and due to the need for an advanced parallel implementation which is independent of the computing hardware, an Alpaka \cite{Matthes17} version was written and its performance was tested.\\

@@

\textcolor{magenta}{
common pre-processing steps:
\begin{itemize}
	\item offset correction \& flat field correction [DSSC \cite{moch2014calibration}, JF \cite{JFcalibration}, EIGER\cite{tinti2018electron} / \cite{dinapoli2013eiger}, MOENCH \cite{cartier2016micrometer}, [MEDIPIX3RX \cite{ballabriga2013medipix3rx}]]
	\item gain correction [DSSC \cite{moch2014calibration}, EIGER \cite{tinti2018electron} / \cite{dinapoli2013eiger}, MOENCH \cite{cartier2016micrometer}]
	\item dynamic gain switching [AGIPD \cite{allahgholi2015agipd}, GOTTHARD \cite{mozzanica2012gotthard}, JF \cite{JFcalibration}, [MEDIPIX3RX \cite{ballabriga2013medipix3rx}]
	\item pixel sorting (on FPGA)[AGPID \cite{becker2013high}]
	\item discard uninteresting events (in readeout circuit)[DSSC \cite{erdinger2012dssc}]
	\item photon counting [MEDIPIX3RX \cite{ballabriga2013medipix3rx}]
	\item charge summing / interpolation (to counter charge sharing; on ASIC)[MEDIPIX3RX \cite{ballabriga2013medipix3rx}]
	\item cluster finder [MOENCH \cite{cartier2016micrometer}]
	\item sub pixel interpolation [MOENCH \cite{cartier2016micrometer}]
\end{itemize}
detector data processing libraries (with pre-processing stage):
\begin{itemize}
	\item Karabo \cite{fangohr2018data}
	\begin{itemize}
		\item developed to support experiments at the European XFEL
		\item written in C++ and Python
		\item pipelined design
		\item supports offline and online operation and an additional rapid-feedback mode
		\item can utilize multiple nodes
		\item also implements file IO and a GUI
		\item extensible through Karabo-Bridge
		\item has OnDA interface
		\item supports Jupytier notebook
	\end{itemize}
	\item Hummingbird \cite{daurer2016hummingbird}
	\begin{itemize}
		\item optimized for real-time usage
		\item implemented in Python
		\item main purpose: analysis of diffraction data at LCLS
		\item achieves rates above 100 Hz
		\item contains a GUI and supports multiple file streams at once
		\item can utilize multiple computing nodes
		\item can read data generated by psana
	\end{itemize}
	\item OnDA \cite{mariani2016onda}
	\begin{itemize}
		\item supports online \& offline operation
		\item can utilize multiple computing nodes (via ZeroMQ and MPI)
		\item master/worker architecture
		\item one program instance per core
		\item designed to give fast online feedback
		\item main objective: real-time monitoring of X-ray imaging experiments
		\item easily adaptable
		\item contains a GUI
		\item implemented in Python
		\item can utilize a psana back end
	\end{itemize}
	\item psana \cite{damiani2016linac}
	\begin{itemize}
		\item = Photon Science ANAlysis
		\item similar functionality to pyana (which is written entirely in Python), but implemented mostly in C++ with a Python interface
		\item used to analyze data produced by SLAC National Laboratory
		\item can utilize multiple compute nodes (via MPI)
		\item supports online \& offline operation
		\item contains detector calibration algorithms (e.g. pedestal calibration, bad-pixel determination, etc.) and GUI front end
		\item designed to give fast online feedback
	\end{itemize}
	\item Cheetah \cite{barty2014cheetah}
	\begin{itemize}
		\item mode for XFEL applications
		\item offline \& online processing
		\item evaluates data quality metrics (e.g. number of Bragg peaks etc.)
		\item retain only frames with high likelihood of being usable
		\item --> main task: data reduction
		\item additional pre-processing
		\item also has a GUI
		\item implemented in C++
	\end{itemize}
	\item CASS \cite{foucar2012cass}
	\begin{itemize}
		\item evaluate data acquired using the CFEL ASG Multi Purpose instrument (CAMP), specifically data collected at FLASH and the European XFEL
		\item supports multiple detectors in online and offline scenarios
		\item implemented in C++ with a Qt front end
		\item consists of input ringbuffer and processing modules (which do pre- \& post-processing)
		\item GUI implemented in two additional programs: joCASSView (plots data) \& luCASSView (interactive)
		\item GUIs communicate with computation nodes via TCP/IP
	\end{itemize}
	\item XDS \cite{kabsch2010xds}
	\begin{itemize}
		\item in development since 1988
		\item mainly deals with crystallography
		\item also does pre-processing steps
		\item no real-time/on-line capability
	\end{itemize}
\end{itemize}
initialized data further used e.g. in crystallography \cite{kabsch2010integration} \cite{white2012crystfel} \cite{grosse2002computational}}\\

@@

\textcolor{magenta}{
\textbf{requirements:}
\begin{itemize}
	\item real-time data processing (at least 1 detector at 2.5 kHz)
	\item simple and extensible design(?)
	\item portable to other architectures (CPU OpenMP, TBB, ...) with minimal code changes
\end{itemize}
optional requirements:
\begin{itemize}
	\item easily adaptable to similar detectors (e.g. MÖNCH)
\end{itemize}
expected bottlenecks:
\begin{itemize}
	\item communication speed between host and GPU
	\item RAM speed
\end{itemize}
\textbf{design overview:}
\begin{itemize}
	\item wrap for most important alpaka functions (for ease of use) --$>$ explain alpaka shortly (?)
	\item templated for maximal performance and configurability (?)
	\item written in modern C++ for optimal performance
	\item pipelined stream processing of data: upload data to GPU -$>$ process -$>$ download again
	\item multiple detectors and multiple GPUs usable in parallel
\end{itemize}
$<$insert data flow diagram here$>$ \\
architecture specific optimizations:
\begin{itemize}
	\item contiguous memory vs. grid striding loop
	\item only upload data if needed
\end{itemize}
\textbf{software architecture:}
\textit{hardware accelerator concept}
\begin{itemize}
	\item host
	\item device
	\item memory not necessarily shared
	\item data might need to be copied onto device
	\item device starts (possibly asynchronous) execution
	\item result is copied back to host
	\item common host: CPU
	\item common devices: GPUs, FPGAs, Multicore CPUs (in general multi- and manycore systems)
\end{itemize}
$<$insert hardware sketch here$>$
\textit{the dispenser}
\begin{itemize}
	\item handles devices (initialization, memory management, etc.)
	\item enqueues data to next free devices (asynchronously)
	\item copies data back from devices (asynchronously)
\end{itemize}
$<$insert dispenser structure here$>$
\textit{the kernel structure}
\begin{itemize}
	\item written in empty struct as static function
	\item all needed values passed to function, no real access to member variables
	\item most data passed by pointer, only single numbers passed directly
	\item kernel has access to device properties (block size, grid size, device specific functions, etc.)
\end{itemize}
$<$insert example here$>$
\textit{easy hardware abstraction}
\begin{itemize}
	\item useful settings defined for every possible device class
	\item device is selected by using statement (similar to a typedef)
	\item no additional changes needed
\end{itemize}
$<$insert example here$>$
}
